Practical Hilbert space approximate Bayesian Gaussian processes for probabilistic programming

### 2022-02-16 Paper/Manuscript_Stats&Comp/manuscript_STCO.tex

\subsection{Theoretical evidence of linear proportionality between $m$, $c$ and $l$} \label{subsec_theoretic_linearity}

There is a clear relation between the number of basis functions $m$ and the boundary factor $c$ with the length-scale $\ell$ of the approximated function. In this section we discuss about the existing theoretical evidence that the minimum number of basis functions $m$ needed for an accurate approximation is linearly proportional to the the boundary factor $c$ and inversely proportional to the lengthscale $l$. In the next Section \ref{subsec_empiric_relations} we conduct an empirical analysis through simulations to prove this expected linear proportionality between $m$, $c$ and $l$ and compute the specific coefficients of proportionality for every kernel.

{\color{blue}Aki: We can also look at the mean zero-level upcrossings, which for QE is
proportional to 1/l which gives as that part, and the effect of c should also
be quite linear when c is not very close to 1 and m is not very small as it
linearly makes the interval shorter so that the there are less zero-level
upcrossings available in the basis functions. You can visualize this using
{\small \url{https://avehtari.github.io/casestudies/Motorcycle/motorcycle_gpcourse.html}} and
look how the plot of 6 basis functions change when you set c=1, c=1.5, c=2.25
(ie $1.5^2$). Instead of looking at mean zero-level crossings, we can look at the
spectral densities and when they go to zero. This is illustrated in the same
notebook after that figure. E.g. with lengthscale l=1,l=1/2, l=1/4, l=1/8 to
explain 99\% of the prior variance we need 4, 8, 16, and 32 basis functions
(linear in l) and to explain 80\% of the prior variance we need 2, 4, 8, 16
basis functions. These spectral densities don't depend on c. When c increaes
the basis functions are truncated and for example going from c=1.5 to c=2.25,
basis function 9 has the flexibility basis function 6 had, so 1.5 more basis
functions are needed. So we have proportionality c/l. }

{\color{red} Gabi: The spectral densities do depend on $c$, see equations (\ref{eq_approxf}) and (\ref{eq_eigenvalue}).}

@@

Empirical simulations are carried out to analyze the relationships between $m$, $c$ and $l$. Figure \ref{fig6_relationships} depicts how these three factors interact and affect the accuracy of the HSGP approximation for a GP with square exponential, Mat\'ern ($\nu$=3/2) and Mat\'ern ($\nu$=5/2) covariance functions and a single input dimension. More precisely, for a given GP model (with a particular covariance function) with length-scale $\ell$ and given a boundary factor $c$, Figure \ref{fig6_relationships} shows the minimum number of basis functions $m$ required to obtain an accurate approximation in the sense of satisfying eq. (\ref{eq_diff_covs}). We considered an approximation to be a sufficiently accurate when the total variation difference between the approximate and exact covariance functions, $\varepsilon$ in eq. (\ref{eq_diff_covs}), is below 1$\%$ of the total area under the curve of the exact covariance function $k$ such that
%
\begin{equation} \label{eq:rel_total_distance}
 \int | k(\tau) - \tilde{k}_m(\tau)|  \,\mathrm{d}\tau  < 0.01 \int k(\tau) \,\mathrm{d}\tau,
\end{equation}
where $\tilde{k}_m$ is the approximate covariance function with $m$ basis functions. Alternatively, these figures can be understood as providing the minimum $c$ that we should use for given $\ell$ and $m$. Of course, we may also read it as providing the minimum $\ell$ that can be approximated with high accuracy given $m$ and $c$. We obtain the following main conclusions:

@@

From the empirical observations, numerical equations governing the relationships between $m$, $c$ and $\ell$ were estimated, which show an agreement with the hypothesis of linear proportionality between $m$, $\frac{l}{S}$ (lenghscale $l$ relative to the half range of the input domain $S$) and $c$. The estimated numerical equations for each particular covariance function are the following:

\paragraph*{Squared exponential}:
%
\begin{align}\label{eq_m_l_QE}
&m \geq 1.75 \, \frac{c}{\nicefrac{l}{S}} \;\; \Leftrightarrow \;\; \nicefrac{l}{S} \geq 1.75 \, \frac{c}{m},
\end{align}
%
with
%
\begin{align}\label{eq_c_vs_l_QE}
&c \geq 3.2 \, \nicefrac{l}{S} \;\; \& \;\, c \geq 1.2      
\end{align}

\paragraph*{Mat\'ern ($\nu$=3/2)}:
%
\begin{align}\label{eq_m_l_mat32}
&m \geq 3.42 \, \frac{c}{\nicefrac{l}{S}} \;\; \Leftrightarrow \;\; \nicefrac{l}{S} \geq 3.42 \, \frac{c}{m},
\end{align}
%
with
%
\begin{align}\label{eq_c_vs_l_mat32}
&c \geq 4.5 \, \nicefrac{l}{S} \;\; \& \;\, c \geq 1.2        
\end{align}

\paragraph*{Mat\'ern ($\nu$=5/2)}:
%
\begin{align}\label{eq_m_l_mat52}
&m \geq 2.65 \, \frac{c}{\nicefrac{l}{S}} \;\; \Leftrightarrow \;\; \nicefrac{l}{S} \geq 2.65 \, \frac{c}{m},
\end{align}
%
with
%
\begin{align}\label{eq_c_vs_l_mat52}
&c \geq 4.1 \, \nicefrac{l}{S} \;\; \& \;\, c \geq 1.2       
\end{align}

Notice that we establish a lower bound at 1.2 for $c$ in every case.

@@

Having obtained an estimate $\hat{\ell}$ of $\ell$ from the HSGP model based on prespecified $m$ and $c$, we can check whether or not $\hat{\ell}$ exceeds the minimum length-scale provided by equations (\ref{eq_m_l_QE}), (\ref{eq_m_l_mat32}) and (\ref{eq_m_l_mat52}) (depending on which kernel is used). If $\hat{\ell}$ exceeds this recommended minimum length-scale, the approximation is assumed to be good. If, however, $\hat{\ell}$ does not exceed recommended minimum length-scale, the approximation may be inaccurate and $m$ and/or $c$ need to be updated. If we look back to the conclusions drawn from Figures~\ref{fig4_MSE_vs_J} and \ref{fig5_lscale_vs_J}, where $m = 10$ basis functions and a boundary factor of $c = 1.5$ were sufficient to obtain an accurate approximation of a function with $\ell = 0.3$, we can recognize that these conclusions also matches those obtained from eq. (\ref{eq_m_l_QE}).

@@

\item[Step A1.] Make an initial guess on the true length-scale $\ell_1$ of the function to be learned. If there is no approximate prior information, we recommend a medium-large length-scale, about $\ell_1 \approx [0.5-1]$. Set the iteration index to $i=1$.

\item[Step A2.] Obtain the minimum valid boundary factor $c_i$ determined by $\ell = \ell_i$ for the given kernel as per Section \ref{sec_num_equations}

>>>>

\subsection{Near linear proportionality between $m$, $c$, and $l$} \label{subsec_theoretic_linearity}

A priori the terms in the series expansion \eqref{eq_approxf_multi} with very small spectral density are unlikely to contribute to the approximation. Given the boundary factor $c$ and the length-scale $\ell$, we can compute cumulative sum of the spectral densities and find out how many basis functions are a priori explaining almost 100\% of the variation. Thus, given $c$ and $\ell$, we can estimate a good choice for the number of basis functions $m$ for any covariance function.

When considering squared exponential and Mat{\'e}rn covariance functions, we can show with simple algebra that when when $c$ is larger than the minimal value recommendation (see TODO ADD REF HERE), the $m$ first terms needed to explain almost 100\% of the variation has near linear relationship with $\ell$ and $c$. With decreasing $\ell$, the $m$ should grow near linearly with $1/\ell$ and with increasing $c$, the $m$ should grow near linearly with $c$.
%
This is natural as with decreasing $\ell$, more higher frequency basis functions are needed. With increasing $c$, the smaller range of the basis functions are used in the approximation, the expected number of zero up-crossings goes down linearly with $c$, and thus more higher frequency basis functions are needed to compensate this.
%
When $c$ is below our recommendations given $\ell$, the effect of $\ell$ and $c$ to the recommended $m$ is more non-linear, but as long as we stay in the recommended range the linearity assumption is useful thumb rule for how to change $m$ if $\ell$ or $c$ is changed.

% There is a clear relation between the number of basis functions $m$ and the boundary factor $c$ with the length-scale $\ell$ of the approximated function. In this section we discuss about the existing theoretical evidence that the minimum number of basis functions $m$ needed for an accurate approximation is linearly proportional to the the boundary factor $c$ and inversely proportional to the lengthscale $l$.
% In the next Section \ref{subsec_empiric_relations} we conduct an empirical analysis through simulations to prove this expected linear proportionality between $m$, $c$ and $l$ and compute the specific coefficients of proportionality for every kernel.

% {\color{blue}Aki: We can also look at the mean zero-level upcrossings, which for QE is
% proportional to 1/l which gives as that part, and the effect of c should also
% be quite linear when c is not very close to 1 and m is not very small as it
% linearly makes the interval shorter so that the there are less zero-level
% upcrossings available in the basis functions. You can visualize this using
% {\small \url{https://avehtari.github.io/casestudies/Motorcycle/motorcycle_gpcourse.html}} and
% look how the plot of 6 basis functions change when you set c=1, c=1.5, c=2.25
% (ie $1.5^2$). Instead of looking at mean zero-level crossings, we can look at the
% spectral densities and when they go to zero. This is illustrated in the same
% notebook after that figure. E.g. with lengthscale l=1,l=1/2, l=1/4, l=1/8 to
% explain 99\% of the prior variance we need 4, 8, 16, and 32 basis functions
% (linear in l) and to explain 80\% of the prior variance we need 2, 4, 8, 16
% basis functions. These spectral densities don't depend on c. When c increaes
% the basis functions are truncated and for example going from c=1.5 to c=2.25,
% basis function 9 has the flexibility basis function 6 had, so 1.5 more basis
% functions are needed. So we have proportionality c/l. }

% {\color{red} Gabi: The spectral densities do depend on $c$, see equations (\ref{eq_approxf}) and (\ref{eq_eigenvalue}).}

@@

Empirical simulations are carried out to analyze the relationships between $m$, $c$ and $l$. Figure \ref{fig6_relationships} depicts how these three factors interact and affect the accuracy of the HSGP approximation for a GP with square exponential, Mat\'ern ($\nu$=3/2) and Mat\'ern ($\nu$=5/2) covariance functions and a single input dimension. More precisely, for a given GP model (with a particular covariance function) with length-scale $\ell$ and given a boundary factor $c$, Figure \ref{fig6_relationships} shows the minimum number of basis functions $m$ required to obtain an accurate approximation in the sense of satisfying eq. (\ref{eq_diff_covs}). We considered an approximation to be a sufficiently accurate when the total variation difference between the approximate and exact covariance functions, $\varepsilon$ in eq. (\ref{eq_diff_covs}), is below 1$\%$ of the total area under the curve of the exact covariance function $k$ such that
%
\begin{equation} \label{eq:rel_total_distance}
 \int | k(\tau) - \tilde{k}_m(\tau)|  \,\mathrm{d}\tau  < 0.01 \int k(\tau) \,\mathrm{d}\tau,
\end{equation}
%
where $\tilde{k}_m$ is the approximate covariance function with $m$ basis functions. Alternatively, these figures can be understood as providing the minimum $c$ that we should use for given $\ell$ and $m$. Of course, we may also read it as providing the minimum $\ell$ that can be approximated with high accuracy given $m$ and $c$. We obtain the following main conclusions:

@@

When $c$ is large enough, there is near
linear proportionality between $m$, $\frac{l}{S}$ (lenghscale $l$ relative to the half range of the input domain $S$) and $c$ (see Section~\ref{subsec_theoretic_linearity}). To obtain practical numerical functions that can be used to guide the selection of these parameters, we have empirically derived the needed constant terms. We establish a lower bound for $c$ set to 1.2.

% From the empirical observations the hypothesis of linear proportionality between $m$, $\frac{l}{S}$ (lenghscale $l$ relative to the half range of the input domain $S$) and $c$ has been proved. Following we show the numerical functions governing these relationships for every kernel.

\paragraph*{Squared exponential}:
%
\begin{align}\label{eq_m_l_QE}
&m= 1.75 \, \frac{c}{\nicefrac{l}{S}} \;\; \Leftrightarrow \;\; \nicefrac{l}{S}= 1.75 \, \frac{c}{m},
\end{align}
%
with
%
\begin{align}\label{eq_c_vs_l_QE}
&c \geq \, 3.2 \, \nicefrac{l}{S} \;\; \& \;\, c \geq \, 1.2      
\end{align}

\paragraph*{Mat\'ern ($\nu$=5/2)}:
%
\begin{align}\label{eq_m_l_mat52}
&m= 2.65 \, \frac{c}{\nicefrac{l}{S}} \;\; \Leftrightarrow \;\; \nicefrac{l}{S}= 2.65 \, \frac{c}{m},
\end{align}
%
with
%
\begin{align}\label{eq_c_vs_l_mat52}
&c \, \geq \, 4.1 \, \nicefrac{l}{S} \;\; \& \;\, c \geq \, 1.2       
\end{align}

\paragraph*{Mat\'ern ($\nu$=3/2)}:
%
\begin{align}\label{eq_m_l_mat32}
&m= 3.42 \, \frac{c}{\nicefrac{l}{S}} \;\; \Leftrightarrow \;\; \nicefrac{l}{S}= 3.42 \, \frac{c}{m},
\end{align}
%
with
%
\begin{align}\label{eq_c_vs_l_mat32}
&c \, \geq \, 4.5 \, \nicefrac{l}{S} \;\; \& \;\, c \geq \, 1.2        
\end{align}

These constants vary monotonically with respect to $\nu$ (squared exponential corresponding to Mat\'ern with $\nu \rightarrow \infty$). Using the formula for Mat\'ern ($\nu$=3/2) provides the largest $m$ and $c$, and thus this formula alone could be used as a conservative choice for all Mat\'ern covariance functions with $\nu \geq 3/2$ and likely as a good initial guess for many other covariance functions. If the aim is to find minimal $m$ to speedup the computation, a further refined formula can be obtained for new covariance functions.

@@

Having obtained an estimate $\hat{\ell}$ of $\ell$ from the HSGP model based on prespecified $m$ and $c$, we can check whether or not $\hat{\ell}$ exceeds the minimum length-scale provided by equations (\ref{eq_m_l_QE}), (\ref{eq_m_l_mat32}) and (\ref{eq_m_l_mat52}) (depending on which kernel is used). If $\hat{\ell}$ exceeds this recommended minimum length-scale, the approximation is assumed to be good. If, however, $\hat{\ell}$ does not exceed recommended minimum length-scale, the approximation may be inaccurate and $m$ and/or $c$ need to be updated. If we look back to the conclusions drawn from Figures~\ref{fig4_MSE_vs_J} and \ref{fig5_lscale_vs_J}, where $m = 10$ basis functions and a boundary factor of $c = 1.5$ were sufficient to obtain an accurate approximation of a function with $\ell = 0.3$, we can recognize that these conclusions also match those obtained from eq. (\ref{eq_m_l_QE}).

@@

\item[Step A1.] Make an initial guess on the lengthscale $l_1$. If there is no useful information available, we recommend to start with a large lengthscale, about $\ell_1 \approx [0.5-1]$. Set the iteration index to $i=1$.

\item[Step A2.] Obtain the minimum valid boundary factor $c_i$ determined by $\ell = \ell_i$ for the given kernel as per Section \ref{sec_num_equations}.
